\section{Exercises}

% 1
\eoce{\qt{The price of pairs}
	As in Section \ref{pairedData}, let $\omega$ be a textbook, let $X(\omega)$ be its price in the UCLA bookstore.
	Let $Y(\omega)$ be its price on Amazon.
	Give an example of sets of observations that are paired and an example of sets of observations that are not paired,
	where an observation is the price of a textbook.
}{}

% 2
\eoce{\qt{Tee of infinity}
	Informally describe the quantity $\underset{n \rightarrow \infty}{\text{lim}}\frac{S}{\sqrt{n}}$, where $S$ is the standard error.
	Conclude that the $t$-distribution approximates the standard distribution as the number of degrees of freedom approaches infinity.
}{}

% 3
\eoce{\qt{Abnormality}
	Briefly describe why each of the following distributions might fail to be normal.
	\begin{enumerate}
		\item 
			200 homes in a 1000-home district are surveyed about their voting habits.
		\item
			10 homes in a 1000-home district are surveyed about their preference for one of two candidates.
	\end{enumerate}
}{}

% 4
\eoce{\qt{Standard error of a difference}
	If $p_1$ and $p_2$ are two population proportions with sample sizes $n_1$ and $n_2$ respectively such that $\hat{p_1}$ and $\hat{p_2}$ are normal,
	verify directly that $SE_{\hat{p_1}-\hat{p_2}} = \sqrt{\frac{p_1(1-p_1)}{n_1} + \frac{p_2(1-p_2)}{n_2}}$.
}{}


\eoce{\qt{Binomial power}
	Let $X$ be a binomial random variable with parameters $n=3$ and $p$. The null hypothesis $H_0$ says that $p=1/3$.
	We decide to reject $H_0$ if the number of heads observed out of $n=3$ is not 1. Find the power function $g(p)$ and graph it.
}{}

\eoce{\qt{Binomial power II}
	Let $X$ be a binomial random variable with parameters $n=4$ and $p$. The null hypothesis $H_0$ says that $p=2/3$.
	We decide to reject $H_0$ if the number of heads observed out of $n=4$ is neither 2 nor 3. Find the power function $g(p)$ and graph it.
}{}