\section{Exercises} \vspace{-2mm}

% 1
\eoce{\qt{One-half factorial\label{factorial}}
	Define the gamma function ${\displaystyle \Gamma(x) := \int_0^{\infty} t^{x-1} e^{-t}dt}$, where $x$ is a nonnegative real number.
	\begin{parts}
		\item
			Show that $\Gamma(x+1) = x \Gamma(x)$ for all $x$. Conclude that $\Gamma(n+1) = n!$ for all natural numbers $n$.
			\textit{Hint: use integration by parts for the first part, and calculate $\Gamma(1)$ and apply induction for the second}.
		\item
			Calculate $\Gamma \left(\frac{1}{2} \right)$.
	\end{parts}
}{}

% 2
\eoce{\qt{Moments with $\chi^2$}
	Let a random variable $X$ have the \textit{chi-square density with $r$ degrees of freedom}, or \textit{$X$ is $\chi^2(r)$} for short, if
	its probability density function is of the form
	\begin{equation}\label{aeq}
		f(x) = \frac{1}{\Gamma \left(\frac{r}{2} \right) 2^{\frac{r}{2}} }x^{\frac{r}{2}-1} e^{-\frac{x}{2}}
	\end{equation}
	\begin{parts}
		\item
			Using Exercise \ref{factorial}(b), show that for any $r$, $f$ from (\ref{aeq}) is indeed a probability density function.
		\item
			If $X = \chi^2(r)$, show that $M_X(t) = \frac{1}{(1-2t)^{\frac{r}{2}}}$, where $M_X(t)$ is the moment generating function of $X$.
		\item
			Use part (b) to show that if $X$ is a normal random variable with expected value 0 and variance 1, then $X^2$ is $\chi^2(1)$.
		\item
			Use part (c) to show that if $X_1, \dots, X_n$ are mutually independent normal random variables,
			each with expected value 0 and variance 1, then $\sum_{i=1}^n X_i^2$ is $\chi^2(n)$.
	\end{parts}
}{}

% 3
\eoce{\qt{Difference of $\chi^2$}
	Let $X_1$ and $X_2$ be independent, let $X_1$ be $\chi^2(r_1)$, and let $X_1+X_2$ be $\chi^2(r)$ for some $r> r_1$.
	Show that $X_2$ is $\chi^2(r-r_1)$.
}{}

% 4
\eoce{\qt{Gender and color preference\label{gender_color_preference_CI_concept}}
	(This is Exercise 6.25 from \emph{OpenIntro Statistics}, copied here because of Exercise \ref{zain_email} below.)
	A 2001
	study asked 1,924 male and 3,666 female undergraduate college students their favorite
	color. A 95\% confidence interval for the difference between the proportions of males
	and females whose favorite color is black $(p_{male} - p_{female})$ was calculated to
	be (0.02, 0.06). Based on this information, determine if the following statements are
	true or false, and explain your reasoning for each statement you identify as
	false.%\footfullcite{Ellis:2001}
	\begin{parts}
		\item We are 95\% confident that the true proportion of males whose favorite color is
		black is 2\% lower to 6\% higher than the true proportion of females whose favorite
		color is black.
		\item We are 95\% confident that the true proportion of males whose favorite color is
		black is 2\% to 6\% higher than the true proportion of females whose favorite color is
		black.
		\item 95\% of random samples will produce 95\% confidence intervals that include the
		true difference between the population proportions of males and females whose favorite
		color is black.
		\item We can conclude that there is a significant difference between the proportions of
		males and females whose favorite color is black and that the difference between the two
		sample proportions is too large to plausibly be due to chance.
		\item The 95\% confidence interval for $(p_{female} - p_{male})$ cannot be calculated
		with only the information given in this exercise.
	\end{parts}
}{}

% 5
\eoce{\qt{Finding $\hat p_x$ and $\hat p_y$\label{zain_email}}
	In Exercise \ref{gender_color_preference_CI_concept}, let $p_y$ be the male proportion and $p_x$ female.
	We need the standard error of $\hat p_y-\hat p_x$ which will be
	$$\sqrt{(\hat p_y(1-\hat p_y)/1924)+(\hat p_x(1-\hat p_x)/3666)}.$$
	From this being 0.01 (since the width of the 95\%, hence $1.96\approx 2$ standard deviations, confidence interval is 0.02) and also $\hat p_y-\hat p_x=0.04$, deduce what they each are.
}{}

% 6
\eoce{\qt{American climate perspectives}
	A 2018 survey\footnote{
		\url{https://ecoamerica.org/wp-content/uploads/2018/10/november-2018-american-climate-perspectives-survey.pdf}
	} found a value $\hat p=79\%$ and reported a margin of error of 3.5\%.
	The sample size was $n=800$ and the confidence level was 95\% as usual. Did they calculate their standard error using $\hat p=.79$ or using an assumed value $p=1/2$? What would the margin of error be with the other choice? %Answer: $1.96\sqrt{.79(1-.79)/800}=2.8\%$ whereas $1.96\sqrt{.5(1-.5)/800}=3.5\%$.
}{}

% 7
\eoce{\qt{Jonas rolls the dice\label{tyler_email}}
	Jonas rolled a die 24 times and obtained:

	\begin{tabular}{l l}
		Number of eyes & Times occurring\\
		1	&	9	\\
		2	&	5	\\
		3	&	6	\\
		4	&	2	\\
		5	&	0	\\
		6	&	2	\\
	\end{tabular}

	Using the $\chi$-square test and \Cref{chiSquareProbabilityTableShort}, determine whether the null hypothesis that the die
	was fair can be rejected at the 5\% level of significance.
	Find $(O_i-E_i)^2/E_i$ for $i=1,2,3,4,5,6$ and find the p-value.
}{}
%
% % 8
\eoce{\qt{Jane rolls the dice}
	Jane rolled a die 24 times and obtained:

	\begin{tabular}{l l}
		Number of eyes & Times occurring\\
		1	&	10	\\
		2	&	5	\\
		3	&	6	\\
		4	&	2	\\
		5	&	1	\\
		6	&	2	\\
	\end{tabular}
	(Compared to Jonas in Exercise \ref{tyler_email}, Jane had a more extreme count for 1 eyes and a less extreme count for 5 eyes.)
	Using the $\chi$-square test, determine whether the null hypothesis that the die
	was fair can be rejected at the 5\% level of significance.
	Find $(O_i-E_i)^2/E_i$ for $i=1,2,3,4,5,6$ and find the p-value.
}{}

\eoce{\qt{Steve's hypothesis}
	Steve rolled a die 24 times and obtained:

	\begin{tabular}{l l}
		Number of eyes & Times occurring\\
		1	&	9	\\
		2	&	5	\\
		3	&	6	\\
		4	&	2	\\
		5	&	0	\\
		6	&	2	\\
	\end{tabular}
	Before starting Steve made the hypothesis that 1 was more likely than the others:
	He hypothesized that $\mathbb P(X=1)=3/8$ and $\mathbb P(X=x)=1/8$ for $x\in\{2,3,4,5,6\}$.

	Using the $\chi$-square test and \Cref{chiSquareProbabilityTableShort}, determine whether Steve's null hypothesis
	can be rejected at the 5\% level of significance.
	Find $(O_i-E_i)^2/E_i$ for $i=1,2,3,4,5,6$ and find the p-value.
}{}

\eoce{\qt{Mary Ann rolls the dice}
Mary Ann rolled a die 24 times and obtained:

	\begin{tabular}{l l}
		Number of eyes & Times occurring\\
		1	&	10	\\
		2	&	5	\\
		3	&	6	\\
		4	&	2	\\
		5	&	1	\\
		6	&	2	\\
	\end{tabular}
	Before starting Mary Ann made the hypothesis that 1 was more likely than the others:
	She hypothesized that $\mathbb P(X=1)=3/8$ and $\mathbb P(X=x)=1/8$ for $x\in\{2,3,4,5,6\}$.

	Using the $\chi$-square test and \Cref{chiSquareProbabilityTableShort}, determine whether Mary Ann's null hypothesis
	can be rejected at the 5\% level of significance.
	Find $(O_i-E_i)^2/E_i$ for $i=1,2,3,4,5,6$ and find the p-value.
}{}
